---
title: "An introduction to calculating FC-1"
author: "W.Nikolaus Prost"
date: "2026-09-24"
output: github_document
editor_options: 
  chunk_output_type: console
---

```{r setup,include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message=F,warning=FALSE)

```

## Update of the OSPAR FC-1

This is a collection of functions to calculate the occurrence of sensitive fish species indicator by OSPAR based on the OSPAR Quality Status Report 2023. The code is based on the idea of assessing the frequency of species' occurrences of in survey hauls as suggested by Probst et al. (2023). This code is prepared for the OSPAR Intermediate Assessment 2027. New features include:

- Besides the binomial integration as described by Probst et al. (2023), the implementation of a new integration method for combining assessment outcomes from multiple surveys based on an average probability of an assessment outcome weighted by the proportion of hauls included.
- The possibility to perform assessments by region (OSPAR regions I-V) or separately by countries within an OSPAR region.
- A function for conducting assessments across multiple assessment periods.


## References

Lynam, C. P., Bluemel, J. K., and Probst, W. N. 2022. Recovery of Sensitive Fish Species. In: The 2023 Quality Status Report for the Northeast Atlantic, 19 pp. OSPAR Commission, London.

OSPAR. 2023. OSPAR Quality status report 2023.

Probst, W. N., Lynam, C. P., Bluemel, J. K., and Clarke, M. 2023. Assessing change in the occurrence of rare species using the binomial distribution. Ecological Indicators, 156.


## To get started

Install the folders 'scripts' and 'spatial data' onto your local computer. You also might need to install obus and dependent packages (e.g. 'DuckDB'):

remotes::install_github("einarhjorleifsson/obus@2a6c1f64ce0fda8c0167888488ae129b29e1e0f6")

For info see ices-tools-prod/icesDatras#32

Then you should be able to use the functions from the folder 'scripts'

```{r,echo=T,results='hide'}
# Load some packages
library(magrittr);library(reshape);library(tidyverse);library(data.table)
library(sf);library(raster);library(terra)
library(ggplot2);library(patchwork);library(ggpubr);library(pals);library(crayon)
library(worrms)     
library(mapplots)
library(obus)

# Source functions ----
source("./scripts/datras.hh.R")
source("./scripts/datras.merge.hhhl.R")
source("./scripts/boa.wrk.hrs.R")
source("./scripts/boa.asmnt.R")
source("./scripts/period.asmnt.R")
source("./scripts/boa.spatial.R")
source("./scripts/ovrvw.spc.R")
source("./scripts/ovrvw.hh.R")

# Load EEZ & OSPAR data
eezs.15<-read_sf("./spatial data/eezs.ospar.regions.shp")
ospar.regs<-read_sf("./spatial data/ospar_regions_simplified.shp")

# Examples of command chain ----
# Get all hauls from all surveys in OSPAR region III
# Need to do this once per region, than can be merged separately for each species
hh.iii.dat<-read_sf("./test data/hh.iii.dat.shp")

# Explore number of hauls and spatial extent
ovrvw.hh.iii<-ovrvw.hh(hh.iii.dat)

# Plot annual number of hauls per survey
ovrvw.hh.iii$n.hauls %>% 
  as.data.frame %>%
  mutate(year=Var1 %>% as.character %>% as.numeric) %>%
  ggplot(aes(x=year,y=Freq,fill=Var2))+
  geom_col(show.legend=F)+
  facet_wrap(.~Var2,scales="free_y")+
  scale_fill_discrete(palette=pals::tol.rainbow)+
  labs(x="Year",y="Number of hauls")


# Plot spatial coverage of hauls per year
ovrvw.hh.iii$spatial.overview
```
```{r, echo=T}
# Merge with abundance/occurrence data for cod
cod.iii<-read.csv("./test data/cod.iii.csv",header=T)
cod.iii[1:3,]

# Assess binomial occurrence (BOA) for cod for all countries in OSPAR Region III
cod.iii.nat<-boa.asmnt(boa.dat=cod.iii,
                       rp=1985:2015,
                       ap=2022:2025,
                       rgnl=F)

cod.iii.nat$integrated.prob.score

# BOA assessment for cod for entire region III
cod.iii.reg<-boa.asmnt(boa.dat=cod.iii,
                       rp=1985:2003,
                       ap=2016:2021,
                       rgnl=T)

# Get an overview on data coverage for cod
ovw.cod.iii<-ovrvw.spc(cod.iii)
ovw.cod.iii$n.hauls
ovw.cod.iii$occ.freq.plot
ovw.cod.iii$spatial.overview
```
```{r,echo=T,message=F}
# Spatial plots
boa.spatial(cod.iii.nat,"bi")

# Assess cod in Region III across multiple periods
cod.iii.per<-period.asmnt(boa.dat=cod.iii,
                         rp=1985:2003,
                         asp=list(2004:2009,2010:2015,2016:2021,2022:2025))

cod.iii.per$summary.plot
```
